---
title: "BMI Prediction with a Neural Network in R"
author: "Fei Qi"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Project Context

This file contains the cleaned portfolio version of my individual contribution to a
four-person AMS 597 Statistical Computing team project at Stony Brook University.

My contribution focused on Research Question 3: predicting BMI with a neural-network
regression model while excluding height and weight from the predictor set.

The original team submission should be retained separately as the historical course
deliverable. This portfolio version isolates Question 3 and corrects the predictor
exclusion logic so that height and weight are explicitly removed before dummy encoding.

# 1. Libraries

```{r}
required <- c(
  "tidyverse",
  "caret",
  "nnet",
  "NeuralNetTools"
)

new_pkgs <- required[!required %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) {
  install.packages(new_pkgs, dependencies = TRUE)
}

library(tidyverse)
library(caret)
library(nnet)
library(NeuralNetTools)
```

# 2. Load Data

Place the source dataset at `data/Final_data.csv`.

```{r}
set.seed(123)

library(here)

data_path <- here("data", "Final_data.csv")

if (!file.exists(data_path)) {
  stop(
    "Dataset not found. Place Final_data.csv in the data/ directory before running."
  )
}

data <- read.csv(data_path, header = TRUE)

cat("Raw dimensions:", dim(data), "\n")
```

# 3. Shared Data Cleaning

The original team project removed leakage variables, redundant engineered features,
and high-cardinality text fields before modeling.

```{r}
drop_cols <- c(
  # Variables derived from / containing Calories_Burned
  "Burns_Calories_Bin",
  "expected_burn",
  "cal_balance",
  "Burns.Calories..per.30.min._bc",

  # Redundant engineered features
  "BMI_calc",
  "cal_from_macros",
  "pct_carbs",
  "protein_per_kg",
  "pct_HRR",
  "pct_maxHR",
  "lean_mass_kg",

  # High-cardinality text variables
  "meal_name",
  "Name.of.Exercise",
  "Benefit",
  "Target.Muscle.Group",
  "Equipment.Needed",
  "Workout"
)

cleaned_data <- data %>%
  select(-all_of(drop_cols))

cleaned_data$Difficulty.Level <- dplyr::recode(
  cleaned_data$Difficulty.Level,
  "Beginner" = 0L,
  "Intermediate" = 1L,
  "Advanced" = 2L
)

cleaned_data$Gender <- ifelse(
  cleaned_data$Gender == "Male",
  1L,
  0L
)

nominal_cats <- c(
  "Workout_Type",
  "meal_type",
  "diet_type",
  "cooking_method",
  "Body.Part",
  "Type.of.Muscle"
)

cleaned_data[nominal_cats] <- lapply(
  cleaned_data[nominal_cats],
  as.factor
)

cat("Cleaned dimensions:", dim(cleaned_data), "\n")
cat("Missing values:", sum(is.na(cleaned_data)), "\n")
cat("Duplicate rows:", sum(duplicated(cleaned_data)), "\n")
```

# 4. Train / Test Split

Because BMI is the response variable for this portfolio analysis, the split is created
with respect to BMI rather than the response used by another team's research question.

```{r}
set.seed(123)

train_idx <- createDataPartition(
  cleaned_data$BMI,
  p = 0.80,
  list = FALSE
)

train_raw <- cleaned_data[train_idx, ]
test_raw  <- cleaned_data[-train_idx, ]

cat("Training rows:", nrow(train_raw), "\n")
cat("Test rows:", nrow(test_raw), "\n")
```

# 5. Exploratory Data Analysis

## BMI by Diet Type

```{r}
ggplot(train_raw, aes(x = diet_type, y = BMI, fill = diet_type)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "BMI Distribution by Diet Type",
    x = "Dietary Category",
    y = "BMI"
  ) +
  theme_minimal() +
  guides(fill = "none")
```

## Fat Percentage vs. BMI

```{r}
ggplot(train_raw, aes(x = Fat_Percentage, y = BMI)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "gam") +
  labs(
    title = "Fat Percentage vs. BMI",
    x = "Body Fat Percentage",
    y = "BMI"
  ) +
  theme_minimal()
```

# 6. Prepare Predictors

BMI is calculated directly from height and weight, so both are explicitly excluded.
`Calories` is also removed because the original analysis found that it behaved as a
strong proxy and produced suspiciously high predictive performance.

```{r}
q3_drop <- c(
  "Height..m.",
  "Weight..kg.",
  "Calories"
)

q3_train <- train_raw %>%
  select(-any_of(q3_drop))

q3_test <- test_raw %>%
  select(-any_of(q3_drop))

stopifnot(!"Height..m." %in% names(q3_train))
stopifnot(!"Weight..kg." %in% names(q3_train))
stopifnot(!"Calories" %in% names(q3_train))
```

## Dummy Encoding

```{r}
dummy_model <- dummyVars(
  BMI ~ .,
  data = q3_train,
  fullRank = TRUE
)

x_train <- predict(
  dummy_model,
  newdata = q3_train
)

x_test <- predict(
  dummy_model,
  newdata = q3_test
)

train_nn <- data.frame(
  BMI = q3_train$BMI,
  x_train
)

test_nn <- data.frame(
  BMI = q3_test$BMI,
  x_test
)

cat("Model predictor count:", ncol(train_nn) - 1, "\n")
```

# 7. Neural Network Tuning

A one-hidden-layer neural network is trained with the `nnet` method through `caret`.
Five-fold cross-validation evaluates combinations of hidden-layer size and weight decay.

The original project trained the neural network on a 2,000-row subset to reduce
computational cost; that design is retained here for comparability.

```{r}
set.seed(123)

subset_idx <- sample(
  seq_len(nrow(train_nn)),
  min(2000, nrow(train_nn))
)

train_subset <- train_nn[subset_idx, ]

ctrl <- trainControl(
  method = "cv",
  number = 5
)

grid <- expand.grid(
  size = c(3, 5, 7),
  decay = c(0.1, 0.5)
)

nn_model <- train(
  BMI ~ .,
  data = train_subset,
  method = "nnet",
  trControl = ctrl,
  tuneGrid = grid,
  preProcess = c("nzv", "center", "scale"),
  linout = TRUE,
  trace = FALSE,
  maxit = 1000,
  MaxNWts = 5000
)

nn_model$bestTune
```

# 8. Predictor Importance

`varImp()` is used to describe relative predictor importance. These scores should not
be interpreted as statistical significance tests.

```{r}
importance <- varImp(nn_model)

plot(
  importance,
  top = 10,
  main = "Top 10 Predictors of BMI"
)
```

# 9. Held-Out Test Evaluation

```{r}
nn_pred <- predict(
  nn_model,
  newdata = test_nn
)

results <- data.frame(
  RMSE = RMSE(nn_pred, test_nn$BMI),
  MAE  = MAE(nn_pred, test_nn$BMI),
  R2   = R2(nn_pred, test_nn$BMI)
)

print(results)
```

## Actual vs. Predicted BMI

```{r}
ggplot(
  data.frame(
    Actual = test_nn$BMI,
    Predicted = nn_pred
  ),
  aes(x = Actual, y = Predicted)
) +
  geom_point(alpha = 0.4) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Neural Network: Predicted vs. Actual BMI",
    subtitle = paste(
      "Test R²:",
      round(results$R2, 3)
    ),
    x = "Actual BMI",
    y = "Predicted BMI"
  ) +
  theme_minimal()
```

# 10. Optional Architecture Visualization

With many dummy-encoded predictors, the fitted network diagram can become visually
dense. It is included for completeness but is not the primary interpretation tool.

```{r}
plotnet(
  nn_model$finalModel,
  max_sp = TRUE
)
```

# 11. Final Model Summary

```{r}
cat("=== FINAL MODEL SUMMARY ===\n")
cat("Method         : Neural Network Regression\n")
cat("Test R²        :", round(results$R2, 3), "\n")
cat(sprintf("Test RMSE      : %.2f BMI units\n", results$RMSE))
cat(sprintf("Test MAE       : %.2f BMI units\n", results$MAE))
```

# Limitations

- Neural-network training uses 2,000 observations rather than the full training set.
- Neural networks are less interpretable than linear regression models.
- Variable-importance scores describe relative contribution, not causal effects or
  statistical significance.
- Fat percentage is a body-composition measurement, so the model should be described as
  excluding height and weight rather than as using only lifestyle and dietary features.
- Results should be regenerated from this cleaned portfolio version before publishing
  updated performance metrics.
